Wildcard: End user modification of web applications using a live data table

### 2020-01-07 paper.tex
in progress

\title{Wildcard: End user modification of web applications through a
data table}
\subtitle{}% optional

@@

% Please always include the abstract.
% The abstract MUST be written according to the directives stated in 
% http://programming-journal.org/submission/
% Failure to adhere to the abstract directives may result in the paper
% being returned to the authors.
\begin{abstract}
Many Web applications do not meet the precise needs of their users.
Browser extensions and user scripts offer a way to customize web
applications, but most people do not have the programming skills to
implement their own customizations.

In this paper, we present a prototype of Wildcard, a live programming
system embedded in the browser that empowers users to casually tweak web
applications without needing to program in Javascript or interact with
the DOM.

Wildcard shows the data from a web page in a table, and maintains a
bidirectional connection between the table and the original page. By
directly manipulating the table, users can perform a wide variety of
modifications: sorting/filtering content, adding private annotations,
using spreadsheet formulas to fetch data from other web services, using
custom UI elements to edit form data, and more.

We present examples of using Wildcard to solve real world problems, and
explain the design principles behind the prototype. In the future, we
envision continuing to build Wildcard into a fully deployed system that
helps make the web into a more malleable medium for all users.
\end{abstract}

@@

This is a familiar situation in a world of web applications that are
frequently updated without user consent. Sometimes there is a browser
extension or user script that fixes an issue, and if the user is a
motivated, programmer they might even be able to implement their own
fix. But for most people in most situations, the only recourse is to
complain to the developers and pray that someone listens---or more
likely, to simply give up. While many have become used to this status
quo, we see it as a tremendous waste of the openness of the Web
platform. In \emph{Personal Dynamic Media} \autocite{kay1977}, Alan Kay
originally envisioned personal computing as a medium that let a user
``mold and channel its power to his own needs,'' but today's software is
more like concrete than clay.

In this paper, we introduce Wildcard, a live programming system embedded
in the browser that enables users to tweak web applications without
programming in Javascript or interacting with the DOM. Wildcard adds a
panel to the bottom of a web page that shows a structured table view of
the main data in the page. The table maintains a bidirectional
connection to the original page---when the user manipulates the table,
the original page gets modified, and vice versa.

In Wildcard, a user can sort Airbnb listings with just one intuitive
click on a table header, with no programming required. Beyond sorting
and filtering data, Wildcard also supports accessing third party APIs,
performing small computations, recording private user annotations, and
using alternate UI widgets to control a page. While Wildcard does not
support any change someone might want to make to a website, it makes a
variety of common and useful changes more accessible to end users, as
well as to programmers who want to avoid tedious low-level programming.

Under the hood, the implementation of Wildcard is straightforward,
because a programmer must manually write an adapter for each website
which uses common web scraping techniques to map the data from the web
page into the Wildcard table format. Although there is still manual
programming involved, the result is very different from traditional
single-use browser extensions, because instead of the programmer
defining a narrow use case, the end user is able to make many different
changes on top of a single site-specific adapter.

In this paper, we present examples of using Wildcard to solve real world
problems, and then explain the system architecture and design principles
behind the prototype:

\begin{itemize}
\tightlist
\item
  \emph{Expose consistent structure}: turning isolated applications into
  a structured medium
\item
  \emph{Encourage casual tweaking}: using frequent software modification
  as a starting point for end user programming
\item
  \emph{Build for an ecosystem}: designing for first party developers,
  third party programmers, and non-programmer end users
\end{itemize}

We also explore related work on web augmentation tools,
spreadsheet-based app builders, and web scraping tools. Our novel
contribution in this paper is to combine ideas from each of these areas
into a new kind of tool: a programming environment which maps data from
web pages onto a live table view, enabling end users to modify the data.

Wildcard is currently an early research prototype, with incomplete
features and limited coverage of sites. We plan to continue privately
testing the system with our own use cases, and then to release the tool
publicly, enabling other programmers to contribute site adapters and
other plugins. Ultimately, we hope that Wildcard makes modifying
websites into a natural part of using the web, as mundane and ubiquitous
as using spreadsheets.

\hypertarget{demos}{%
\section{Demos}\label{demos}}

To get a sense of how it feels to use Wildcard, let's see an example of
using it to help with booking a trip using the travel sites Airbnb and
Expedia.

\hypertarget{sorting-and-filtering}{%
\subsection{Sorting and filtering}\label{sorting-and-filtering}}

We start by opening up the Airbnb search listings page to look for a
place to stay. As mentioned before, this page doesn't let us sort by
price, but we can use Wildcard to fix that. First, we open up the
Wildcard panel, which shows a table corresponding to the search results
in the page. As we click around in the table, the corresponding row in
the page is highlighted so we can see the connection between the views.

When we click the Price column header in the table, both the table and
the original page become sorted by price. We can also filter listings by
rating, another feature not offered in the Airbnb site.

After finishing the sorting and filtering, we can close the table view
and continue using the website in its original polished design.

\hypertarget{row-actions}{%
\subsection{Row actions}\label{row-actions}}

@@

Within the site adapter, each action is implemented as a regular
Javascript function running in the context of the page, which can click
on buttons in the UI, launch AJAX requests, navigate to a new page, etc.

\hypertarget{user-annotations}{%
\subsection{User annotations}\label{user-annotations}}

It's a common practice in spreadsheets to add a column next to a table
for recording notes about the item in each row. In Wildcard, this
pattern can be used to create private annotations in a website. Here, we
use this feature to jot down pros and cons of various listings:

In order to match the annotations with the existing design, the site
adapter specifies a reasonable location in each row for injecting new
data.

Annotations are persisted in browser local storage so the user can come
back and view them later. The site adapter associates each table row
with a stable identifier from the original site so that annotations can
be displayed with the appropriate item.

Currently annotations are private to the browser of the user that
created them, but it could be useful in the future to allow users to
share annotations.

\hypertarget{computation-with-formulas}{%
\subsection{Computation with formulas}\label{computation-with-formulas}}

The demos so far have shown straightforward tweaks that provide useful
conveniences with little effort. But Wildcard also supports more
sophisticated tweaks that fetch external data and perform computations,
using a formula language.

When traveling without a car, it's nice to evaluate potential places to
stay based on how walkable the surroundings. Using Wildcard formulas, we
can integrate Airbnb with Walkscore, an API that can rate the
walkability of any location on a 1-100 scale. When we call the
\texttt{WALKSCORE} formula with the latitude and longitude of the
listing, it returns the score as the cell value. Because the cell is
shown in the page, the score also shows up in the page body.

Wildcard formulas are just Javascript functions that take table data as
input and return new data. Formulas can have limited side effects like
accessing external APIs to fetch data, but do not directly manipulate
the page. Instead, they return data into the table, which is then
injected into the page with the same mechanism used for injecting user
annotations.

\hypertarget{custom-ui-elements}{%
\subsection{Custom UI elements}\label{custom-ui-elements}}

@@

In previous examples the data extracted from the site was marked as
read-only (it's not allowed to change the name of an Airbnb listing),
but in this case, the cells are marked as writable, so that changes in
the table are reflected in the original form. This becomes particularly
useful when combined with custom cell editing UIs, which allow users to
graphically edit the value of a cell.

Here's an example of using those features to help with filling in a
form. Filling in dates for a flight search typically requires opening up
a separate calendar app to look up the correct dates, and then manually
copying them into the form. In Wildcard, we can make this easier by
editing the date cell using a datepicker widget that has privileged
access to our calendar information.

Custom UIs enable people to use a consistent UI to enter common types of
data across the web. They also allow a user to access their own private
data as part of a web interface, without needing to expose it to the
website server.

Overall, an interactive data table is a computation model that presents
a surprisingly large range of useful possibilities for end user
modification of websites, while also remaining familiar and easy to use.
While we've presented concretely illustrated here just a few of these
possibilities, we plan to continue developing site adapters, formulas,
and UI elements to explore more use cases, and to eventually publicly
release the tool to allow real end users to discover their own uses.

@@

For extracting data from the page and getting it into structured form,
Wildcard provides ways to concisely and declaratively express scraping
logic. For example, here is a code snippet for extracting the name of an
Airbnb listing. It specifies the field name and some metadata
properties, and then a function that extracts the name div from a row
div.

@@

The site adapter also needs to support the reverse direction: sending
updates from the table to the original page. Most DOM manipulation is
not performed directly by the site adapter: the adapter specifies how to
find the divs representing data rows, and the core platform mutates the
DOM to reflect the table state. The only exception is site-specific
actions (like favoriting in Airbnb) which runs Javascript code that can
mutate the DOM or perform other arbitrary behaviors.

So far we have only implemented three site adapters in Wildcard, but we
plan to implement many more in the near future to continue to test the
limits of the API. Eventually we will also encourage third party
programmers to build and maintain adapters for popular sites.

@@

The design of Wildcard is grounded in several principles, informed by
prior work and our own experimentation. We think these principles extend
beyond this specific system and can more generally inform the design of
end user programming tools, particularly systems enabling users to
modify GUI apps.

@@

In \emph{Changing Minds} \autocite{disessa2000}, Andrea diSessa
critiques the design of modern software with a story about a
hypothetical ``nightmare bike.'' Each gear on the nightmare bike is
labeled not with a number, but with an icon describing its intended use:
smooth pavement uphill, smooth pavement downhill, gravel, etc. This
might seem more ``user-friendly'' than numbered gears, but in fact, it
makes it harder to operate the bike. An ordered numerical sequence
allows the user to intuitively grasp the structure of the system---``if
I shift up, it gets harder to pedal''---but isolated modes provide a
superficial understanding with no grounding in structure. This
understanding might be sufficient for the most common cases, but breaks
down in unfamiliar situations. If someone needs to go uphill on gravel,
do they need to try every mode at random?

Today, most personal computing consists of using distinct apps, each one
an isolated silo of concepts, interfaces, data, and functionality. While
there are some shared idioms and limited points of interoperability, the
fundamental expectation is that each app operates independently. If a
new need emerges, a user must hope ``there's an app for that''; creating
a new tool to meet the need is often out of the question.

Computing does not need to operate this way. As one example, UNIX offers
a compelling alternate vision of small tools connected through a common
abstraction of text files. Just like riding a bike with numbered gears,
UNIX requires users to build up a structured understanding of the
system, which then allows them to flexibly react to new needs not served
by existing tools. Users can also deeply master a text editor and use it
for a vast array of tasks, even as an interactive input mechanism in
shell programs (e.g.~for editing git commit messages). Beaudouin-Lafon
and Mackay call this style of interaction \emph{polymorphic}
\autocite{beaudouin-lafon2000}, since a single interface can be used in
many contexts. diSessa \autocite{disessa2000} notes the deep connection
between polymorphism and literacy: textual literacy rests on a single
rich medium of writing, which once mastered, can be adapted to many
different genres and uses. If people needed to relearn from scratch when
switching from writing essays to writing emails, the medium would lose
most of its potency; this is the situation in software today.

With Wildcard, we hope to make isolated Web applications work a little
bit more like UNIX, by imposing a consistent and simple structure onto
many isolated applications. Just as UNIX uses text files to represent
all data, Wildcard uses relational tables. This abstraction strikes a
balance between being simple and generic: a data table is much simpler
than the DOM tree that describes all the details of the UI, but is also
generic enough to describe the essence of many different interfaces.

@@

Exposing more structured data to users provides improved quality and
consistency of user interfaces. When UI widgets can compete on their
merits rather than having a monopoly on a given data silo or
application, it creates stronger competition at the interface level. For
example, there is competition among email clients which consume an open
protocol, but not among Facebook or Twitter clients. Exposing structure
also leads to improved UI consistency across apps: UNIX users can reuse
their text editor everywhere, but Web users must use a different rich
text editor for each site. Tim Berners-Lee has written in more detail
about these benefits \autocite{berners-lee2018} in the context of the
SOLID project, which envisions user-controlled data as a mechanism for
decoupling data from interfaces, e.g.~giving users a choice of which
client to use to consume a given social media feed. Wildcard has
overlapping goals with SOLID, but does not require decentralized user
control of data---the data can remain on a centralized server, as long
as the interface can be tweaked by end users.

Layering a structured view on top of an existing application is not the
only option available. Systems like Chickenfoot \autocite{bolin2005} and
CoScripter \autocite{leshed2008} forego the intermediate layer of
structure, instead allowing users to create scripts in an informal
language and then perform fuzzy pattern matching to find elements in the
DOM. For example, to find a button after a textbox in Chickenfoot, the
user could type \texttt{click(find(“button\ just\ after\ textbox”))}.
These designs allow for expressing a wide range of operations, but they
don't explicitly indicate what operations are possible---the user can
only see the original page and imagine the possibilities. In contrast,
Wildcard provides affordances that clearly suggest the availability of
certain actions (e.g.~sorting, editing a cell, adding a column with a
derived value), especially to users who are familiar with spreadsheets.
In addition to giving users more certainty about whether a modification
is possible, these affordances might improve discoverability and give
users new ideas for things to try. We also think that sloppy programming
does little to solve the problem of thinking of applications as isolated
worlds, whereas Wildcard contributes to giving users a structured
understanding of the different apps that they use.

@@

\hypertarget{encourage-casual-tweaking}{%
\subsection{Encourage casual tweaking}\label{encourage-casual-tweaking}}

In order to serve the widest possible audience, Wildcard is designed for
tweaking existing software for one's own needs, not for designing a new
application from scratch. Creating a new application requires motivation
and other skills besides programming that not everyone has, but having
frustrations with some aspect of an app which could be solved by a small
tweak is a nearly universal experience.

\begin{longtable}[]{@{}lll@{}}
\toprule
\endhead
& \emph{Casual} & \emph{Not casual}\tabularnewline
\emph{End user friendly} & \textbf{Wildcard} & IFTTT\tabularnewline
\emph{Requires programming} & browser dev tools & editing open source
software\tabularnewline
\bottomrule
\end{longtable}

We can evaluate a system for tweaking software along two dimensions.
First, the technical capability required of the user: is programming
knowledge needed? Second, the level of effort required: how far out of
their way the user must go to make a change? Can they casually make a
small change, or do they need to make a larger project out of it? These
dimensions are not orthogonal, but they are distinct. For example,
setting up a workflow trigger in an end user programming system like
\href{https://ifttt.com/}{IFTTT} does not require much technical skill,
but it does require leaving the user's normal software and entering a
separate environment. On the other hand, running a Javascript snippet in
the browser console requires programming skills, but can be done
immediately and casually in the flow of using a website.

In addition to requiring no programming skills, Wildcard also aims to
support frequent, small modifications. The Wildcard table appears in the
course of normal web browsing, to ensure that the tools for modification
are close at hand while using the original software. Ink and Switch
refers to this property as having an ``in-place toolchain''
\autocite{inkandswitch2019}. Wildcard also provides live feedback,
immediately editing the page in response to user interaction, which
gives users confidence in the changes they are making.

Wildcard also aims to make some valuable changes possible with
particularly low effort: for example, sorting a table in a single click,
or adding an annotation by filling in a cell. We would expect many
Wildcard users to start by using these bits of functionality, before
possibly moving on to more complex features like formulas. This property
is inspired by spreadsheets, which can be used productively even by
someone who has learned only a tiny part of their functionality,
e.g.~for storing tables of numbers or computing simple sums
\autocite{nardi1991}. In contrast, many traditional programming systems
require someone to spend days learning complex concepts before being
able to create a useful piece of software---printing ``hello world'' is
not a genuinely valuable activity.

\hypertarget{build-for-an-ecosystem}{%
\subsection{Build for an ecosystem}\label{build-for-an-ecosystem}}

@@

This approach takes advantage of the unusual openness of the Web
platform. On many other platforms (e.g.~smartphone operating systems),
software is locked down unless first-party developers explicitly provide
hooks for plugins and interoperability, but on the Web, all client-side
code is available for browser extensions to modify. Application authors
can use practices that make it easier to modify their apps (e.g.~clean
semantic markup), or more difficult (e.g.~code obfuscation), but the
default state is openness. This gives extensions freedom to modify
applications in creative ways that the original developers did not plan
for. At the most extreme, this can even include modifications that are
hostile to the desires of the original developers, a notion Cory
Doctorow calls
\href{https://www.eff.org/deeplinks/2019/10/adversarial-interoperability/}{adversarial
interoperability}. Although ad blocking extensions make this style of
extension very prominent, most ideas for extending software merely fall
outside the range of possibilities originally imagined by the developer,
rather than being actively hostile.

Wildcard takes advantage of this openness, and does not depend on
cooperation from first-party website developers. Any programmer can add
support for any website to Wildcard by building a third party adapter.
However, we also hope that eventually first party website developers
will build in Wildcard support to their applications, since this would
reduce the burden of maintaining adapters and make Wildcard plugins more
stable. Implementing the Wildcard adapter API could help developers by
allowing users to fix some of their own issues, and would not
necessarily require too much effort---we envision making it
straightforward to add Wildcard support in a client-side application
which already has access to the structured data in the page. There is
also precedent for first parties implementing an official client
extension API in response to user demand: for several years, Google
maintained an official extension API in Gmail for Greasemonkey scripts
to use. (Incidentally, since then, third parties have continued to
maintain stable Gmail extension APIs used by many browser extensions
\autocite{streak,talwar2019}, illustrating the potential of
collaboratively maintaining third party adapters.)

@@

While similar in high-level goals, Wildcard employs a different solution
strategy from these projects: whereas they generally require building
new software from scratch for that environment, Wildcard instead aims to
maximize the malleability of already existing software. This approach
has the pragmatic benefit of immediately being useful in a much broader
context, although it also requires working within rigid constraints. We
also see Wildcard growing beyond its current bounds in the future and
becoming more similar to these other systems. With substantial future
work, Wildcard could grow from merely being a platform for tweaking
existing software into a platform for building new software from
scratch, designed from the ground up for end user tweakability.

@@

Wildcard's goals are closely shared with other systems which provide
interfaces in the browser for end users to augment and modify websites
while using them.

@@

It is a powerful realization to notice that a spreadsheet can serve as
an end-user-friendly backing data store and computation layer for an
interactive web application. Research projects like Quilt
\autocite{benson2014}, Gneiss \autocite{chang2014}, and Marmite
\autocite{wong2007}, as well as commercial tools like Airtable Blocks
\autocite{zotero-79} and Glide \autocite{zotero-81} allow users to view
data in a spreadsheet table, compute over the data using formulas, and
connect the table to a GUI. Because many users are already familiar with
using spreadsheets, this way of creating applications tends to be much
easier than traditional software methods; for example, in a user study
of Quilt, many people were able to create applications in under 10
minutes, even if they expected it would take them many hours.

Wildcard builds on this powerful idea, but applies it to modifying
existing applications, rather than building new applications from
scratch. For many people, we suspect that tweaking existing applications
provides more motivation as a starting point for programming than
creating a new application from scratch.

>>>>

\title{Wildcard: End user modification of web applications using a live
data table}
\subtitle{}% optional

@@

% Please always include the abstract.
% The abstract MUST be written according to the directives stated in 
% http://programming-journal.org/submission/
% Failure to adhere to the abstract directives may result in the paper
% being returned to the authors.
\begin{abstract}
Many Web applications do not meet the particular needs of their users.
Browser extensions and user scripts offer a way to customize web
applications, but most people do not have the programming skills to
implement their own customizations.

In this paper, we present a prototype of Wildcard, an extension
framework embedded in the browser that empowers users to casually tweak
web applications without needing to program in Javascript or directly
interact with the DOM.

Wildcard shows the data from a web page in a table, and maintains a live
bidirectional connection between the table and the original page. By
directly manipulating the table, users can perform a wide variety of
modifications: sorting/filtering content, adding private annotations,
using spreadsheet formulas to fetch data from other web services, adding
custom UI elements to edit form data, and more.

We present examples of using Wildcard to solve real world problems, and
explain the design principles behind the prototype. In the future, we
envision growing Wildcard into a fully deployed system that helps make
the web into a more malleable medium for all its users.
\end{abstract}

@@

This is a familiar situation in a world of web applications that are
frequently updated without user consent. Sometimes there is a browser
extension or user script that fixes an issue, and if the user is both
motivated and skilled they might even be able to implement their own
fix. But for most people in most situations, the only recourse is to
complain to the developers and pray that someone listens---or more
likely, to simply give up. While many have become used to this status
quo, we see it as a tremendous waste of the openness of the Web
platform. Back in 1977, in \emph{Personal Dynamic Media}
\autocite{kay1977}, Alan Kay had originally envisioned personal
computing as a medium that let a user ``mold and channel its power to
his own needs,'' but today's software is more like concrete than clay.

In this paper, we introduce Wildcard, an extension framework embedded in
the browser that lets users tweak web applications without programming
in Javascript or interacting with the DOM. Wildcard shows a structured
table view of the main data in a webpage, which maintains a live
bidirectional connection to the original UI: when the user manipulates
the table, the UI is instantly modified, and vice versa. This simple
framework allows a wide range of possible modifications, ranging from
sorting lists to fetching related data from other websites.

The system builds on two insights from prior work. The first insight is
that a spreadsheet can serve as the backend for a GUI application, which
enables an end user to build applications in an extremely accessible
environment without doing traditional programming. The second insight is
that web scraping---extracting structured data from an unstructured web
page---unlocks value by making it possible to compute over that data.
Our contribution in this work is to combine those two ideas in a novel
way. By extracting structured data from an application and showing that
data in a live spreadsheet table, Wildcard enables a wide class of end
user modifications to existing applications. Wildcard makes it seem like
an application is backed by a spreadsheet, even though the actual
implementation is not. This allows the user to easily manipulate and
compute over the application's underlying data just as they would in a
spreadsheet, and then to immediately see the results of their changes in
the original application's UI.

Wildcard embodies several principles we propose for designing systems
that enable end user modification of existing software:

\emph{Expose consistent structure}: Most applications only expose a user
interface, without showing the structure of the data underneath.
Furthermore, every application has a different interface, requiring
users to learn many ways of using apps. Wildcard shows data from diverse
applications in a single consistent structure, so that users are able to
invest in learning one mechanism for manipulating the data in their
applications, and for programming with that data.

\emph{Low floor, high ceiling}: We aim to provide a low barrier to entry
so that non-programmers can casually use the system, but a high ceiling
on the complexity it can achieve. Small but genuinely valuable tweaks
like sorting a page are made possible with little effort in the flow of
normal use to provide a low floor, and a rich formula system creates a
high ceiling on possible modifications.

\emph{Build for multiple tiers of users}: We aim to make Wildcard as
easy as possible for end users, by designing roles for tiers of users
with differing levels of technical sophistication. Programmers or
tech-savvy end users are responsible for the data extraction step of
mapping existing UIs to structured data. Most end users only interact
with the structured data, providing a predictable and user-friendly
experience.

Wildcard is currently an early research prototype, with incomplete
features and limited coverage of sites. We plan to continue privately
testing the system with our own use cases, and then to release the tool
publicly, enabling other programmers to contribute site adapters and
other plugins. Ultimately, we hope that Wildcard makes modifying
websites into a natural part of using the web, as ubiquitous as using
spreadsheets in everyday computing.

\hypertarget{demo-booking-a-trip-with-wildcard}{%
\section{Demo: booking a trip with
Wildcard}\label{demo-booking-a-trip-with-wildcard}}

To get a sense of how it might feel to use Wildcard, let's see an
example of someone using it to help with booking a trip on the travel
sites Airbnb and Expedia.

The user starts by opening up the Airbnb search listings page to look
for a place to stay. The page looks nice and mostly works well, but is
missing some key features. As mentioned before, this page doesn't allow
the user to sort by price. It also doesn't let them filter by user
rating. Using Wildcard, the user can add these small features, while
leaving the page's design and the rest of its functionality unchanged.

First, the user opens up the Wildcard panel, which shows a table
corresponding to the search results in the page. As they click around in
the table, the corresponding row in the page is highlighted so they can
see the connection between the views.

Then, the user can use standard spreadsheet column header features to
sort the page by price and filter by rating:

Notice how after manipulating the data, the user was able to close the
table view and continue using the website with its original visual
design. The table view offers a way to change the data backing a page,
but does not need to replace the original interface entirely.

@@

Now the user would like to jot down some notes on the pros and cons of
each listing. To do this, they can simply type notes into an additional
column next to each listing, and the notes appear inside the listings in
the original UI. These annotations are also persisted in the browser for
future retrieval.

Wildcard also includes a formula language which enables more
sophisticated tweaks that fetch external data and perform computations.

When traveling without a car, it's nice to evaluate potential places to
stay based on how walkable the surroundings. Using Wildcard formulas, we
can integrate Airbnb with Walkscore, an API that rates the walkability
of any location on a 1-100 scale. When we call the \texttt{WALKSCORE}
formula with the latitude and longitude of the listing, it returns the
walk score as the cell value. Because the cell's contents are injected
into the page, the score also shows up in the page body.

@@

In previous examples the data extracted from the site was marked as
read-only; users cannot change the name or price of an Airbnb listing.
In this next case, the cells are marked as writable, so that changes in
the table are reflected in the form inputs. This becomes useful when
combined with GUI widgets for editing the value of a table cell.

Filling in dates for a flight search typically requires opening up a
separate calendar app to find the right dates, and then manually copying
them into the form. In Wildcard, we can make this easier by providing a
datepicker widget that has privileged access to the user's calendar
information.

Here we've presented just a few possibilities for how to use Wildcard.
We think the interactive data table offers a flexible computational
model that can support a wide range of other useful modifications, all
while remaining familiar and easy to use.

@@

For extracting data from the page and getting it into structured form,
Wildcard provides ways to concisely and declaratively express scraping
logic. For example, here is a code snippet for extracting the name of an
Airbnb listing. It specifies the field name and some metadata
properties, and then a function that, given the entire listing, returns
the DOM element representing the name of the listing.

@@

The site adapter also needs to support the reverse direction: sending
updates from the table to the original page. Most DOM manipulation is
not performed directly by the site adapter: the adapter specifies how to
find the divs representing data rows, and the core platform mutates the
DOM to reflect the table state. The only exception is site-specific
actions (like favoriting in Airbnb): actions are implemented as regular
Javascript functions running in the context of the page, which can
mutate the DOM, simulate clicks on buttons in the UI, launch AJAX
requests, navigate to new pages, etc.

@@

The design of Wildcard is grounded in several principles, informed by
prior work and our own experimentation. We think these principles extend
beyond Wildcard and might inform the design of other end user
programming tools, particularly systems enabling users to modify GUI
apps.

@@

Today, most personal computing consists of using distinct apps, each one
an isolated silo of data and functionality. While there are some shared
idioms and limited points of interoperability, apps fundamentally
operate independently from one another. Computing does not need to
operate this way---for example, UNIX offers a compelling alternate
vision: many small tools, all built on the consistent abstraction of
text files. Users can invest in deeply mastering a text editor, and then
use it for a vast array of tasks, even as an interactive input mechanism
in shell programs (e.g.~for editing git commit messages).
Beaudouin-Lafon and Mackay call this style of interaction
\emph{polymorphic} \autocite{beaudouin-lafon2000}, since a single
interface widget can be used in many contexts. diSessa
\autocite{disessa2000} has also noted the deep connection between
polymorphism and literacy: textual literacy rests on a single rich
medium of writing, which once mastered, can be adapted to many different
genres and uses. If people needed to relearn from scratch when switching
from writing essays to writing emails, the medium would lose most of its
potency.

With Wildcard, we hope to make isolated Web applications work more like
UNIX, by imposing a consistent and simple data structure onto many
applications. Just as UNIX uses text files to represent all data,
Wildcard uses relational tables. This abstraction strikes a balance
between being simple and generic: a data table is much simpler than the
DOM tree that describes all the details of the UI, but is also generic
enough to describe the essence of many different interfaces.

@@

Exposing structure gives users more choice When UI widgets can compete
on their merits rather than having a monopoly on a given data silo or
application, it creates stronger competition at the interface level. For
example, there is competition among email clients which consume an open
protocol, but not among Facebook or Twitter clients. Exposing structure
also leads to improved UI consistency across apps: UNIX users can reuse
their text editor everywhere, but Web users must use a different rich
text editor for each site. Tim Berners-Lee has written in more detail
about these benefits \autocite{berners-lee2018} in the context of the
SOLID project, which envisions giving users control of their own data as
a mechanism for decoupling data from interfaces (e.g., giving users a
choice of which client to use to consume a given social media feed).
Wildcard has overlapping goals with SOLID, but does not require
decentralized user control of data---the data can remain on a
centralized server, as long as the interface can be tweaked by end
users.

\emph{todo: move the sloppy contrast, and SOLID, down into related work}

Layering a structured view on top of an existing application is not the
only option available. ``Sloppy programming'' systems like Chickenfoot
\autocite{bolin2005} and CoScripter \autocite{leshed2008} forego the
intermediate layer of structure, instead allowing users to create
scripts in an informal language and then perform fuzzy pattern matching
to find elements in the DOM. For example, to find a button after a
textbox in Chickenfoot, the user could type
\texttt{click(find(“button\ just\ after\ textbox”))}. These designs
allow for expressing a wide range of operations, but they don't
explicitly indicate what operations are possible---the user can only see
the original page and imagine the possibilities. In contrast, Wildcard
provides affordances that clearly suggest the availability of certain
actions (e.g.~sorting, editing a cell, adding a column with a derived
value), especially to users who are familiar with spreadsheets. In
addition to giving users more certainty about whether a modification is
possible, these affordances might improve discoverability and give users
new ideas for things to try. ``Sloppy programming'' does give end-users
greater access to modify the apps that they use, but does not seem to
provide them with a deeper understanding of those apps or how they
relate to each other, so the apps are more likely to remain siloed.

@@

\hypertarget{low-floor-high-ceiling}{%
\subsection{Low floor, high ceiling}\label{low-floor-high-ceiling}}

Seymour Papert advocated for programming systems with a ``low floor''
and a ``high ceiling''---making it easy for novices to get started, but
also providing possible a large range of possibilities for more
sophisticated users \autocite{resnick2016}. Unfortunately, many
programming systems only provide either one or the other
\autocite{myers2000}.

\begin{itemize}
\tightlist
\item
  tweaking vs creation from scratch
\item
  in-place toolchain
\item
  easy things easy
\item
  easy things valuable
\item
  this seems clearly justified by the demos
\end{itemize}

rewrite\ldots{}

\begin{longtable}[]{@{}llc@{}}
\toprule
\endhead
& \emph{In-place} & \emph{Not in-place}\tabularnewline
\emph{End user friendly} & \textbf{Wildcard} & IFTTT\tabularnewline
\emph{Requires programming} & browser dev tools & editing open source
software\tabularnewline
\bottomrule
\end{longtable}

We can evaluate a system for tweaking software along two dimensions.
First, the technical capability required of the user: is programming
knowledge needed? Second, whether the change is ``in-place''
\autocite{inkandswitch2019}: can it be made within the same environment
that the software is typically used in? These dimensions are not
orthogonal, but they are distinct. For example, setting up a workflow
trigger in an end user programming system like
\href{https://ifttt.com/}{IFTTT} does not require much technical skill,
but it does require leaving the user's normal software and entering a
separate environment. On the other hand, running a Javascript snippet in
the browser console requires programming skills, but can be done
immediately and casually in the flow of using a website.

In addition to requiring no programming skills, Wildcard supports
frequent, small modifications. The Wildcard table appears in the course
of normal web browsing, to ensure that the tools for modification are
close at hand while using the original software.Wildcard also provides
live feedback, immediately editing the page in response to user
interaction, which gives users confidence in the changes they are
making.

Spreadsheets can be used in genuinely useful ways even by someone who
has learned only a tiny part of their functionality, e.g.~for storing
tables of numbers or computing simple sums \autocite{nardi1991}. This
creates inherent motivation to continue using the tool and eventually
learn its more powerful features. In contrast, many traditional
programming systems require an enormous time investment before a
beginner is able to create any valuable software. Like spreadsheets,
Wildcard aims to support genuinely useful actions with little effort:
for example, sorting a table with a single click, or adding an
annotation by typing into a cell. We would expect many Wildcard users to
start by using these bits of functionality, before possibly moving on to
more complex features like formulas.

High ceiling?

\hypertarget{build-for-multiple-tiers-of-users}{%
\subsection{Build for multiple tiers of
users}\label{build-for-multiple-tiers-of-users}}

\begin{itemize}
\tightlist
\item
  make things easy for the last tier of users (low floor)
\item
  expose predictable structure, don't make them participate in data
  extraction
\end{itemize}

@@

This approach takes advantage of the unusual openness of the Web
platform. On many other platforms (e.g.~smartphone operating systems),
software is locked down unless first-party developers explicitly provide
hooks for plugins and interoperability, but on the Web, all client-side
code is available for browser extensions to modify. Application authors
can use practices that make it easier to modify their apps (e.g.~clean
semantic markup), or more difficult (e.g.~code obfuscation), but the
default state is openness. This gives extensions freedom to modify
applications in creative ways that the original developers did not plan
for. At the most extreme, this can even include modifications that are
hostile to the desires of the original developers. Although ad blocking
extensions make this style of extension very prominent, most ideas for
extending software merely fall outside the range of possibilities
originally imagined by the developer, rather than being actively
hostile.

Wildcard takes advantage of this openness, and does not depend on
cooperation from first-party website developers: any programmer can add
support for any website to Wildcard by building a third party adapter.
But the platform is still open to first parties developing their own
adapters, which would often be in their interests. Providing Wildcard
support would allow users to build extensions to fulfill their own
feature requests. It also would not necessarily require much effort:
adding Wildcard support would be more straightforward for a first-party
than a third-party because they have direct access to the structured
data in the page. There is also precedent for first parties implementing
an official client extension API in response to user demand: for several
years, Google maintained an official extension API in Gmail for
Greasemonkey scripts to use. (Incidentally, since then, third parties
have continued to maintain stable Gmail extension APIs used by many
browser extensions \autocite{streak,talwar2019}, illustrating the
potential of collaboratively maintaining third party adapters.)

@@

While similar in high-level goals, Wildcard employs a different solution
strategy from these projects: whereas they generally require building
new software from scratch for that environment, Wildcard instead aims to
maximize the malleability of already existing software. This approach
has the pragmatic benefit of immediately being useful in a much broader
context, although it also requires working within rigid constraints. We
also see Wildcard growing beyond its current bounds and becoming more
similar in the future to these other systems. With substantial future
work, Wildcard could grow from merely being a platform for tweaking
existing software into a platform for building new software from
scratch, designed from the ground up for end user tweakability.

@@

Wildcard's goals are closely shared with other systems that provide
interfaces in the browser for end users to augment and modify websites
while using them.

@@

Prior work has made the powerful realization that a spreadsheet can
serve as an end-user-friendly backing data store and computation layer
for an interactive web application. Research projects like Object
Spreadsheets \autocite{mccutchen2016}, Quilt \autocite{benson2014},
Gneiss \autocite{chang2014}, and Marmite \autocite{wong2007}, as well as
commercial tools like Airtable Blocks \autocite{zotero-79} and Glide
\autocite{zotero-81} allow users to view data in a spreadsheet table,
compute over the data using formulas, and connect the table to a GUI.
Because many users are already familiar with using spreadsheets, this
way of creating applications tends to be far easier than traditional
software methods; for example, in a user study of Quilt, many people
were able to create applications in under 10 minutes, even if they
expected it would take them many hours.

Wildcard builds on this idea, but applies it to modifying existing
applications, rather than building new applications from scratch. For
many people, we suspect that tweaking existing applications provides
more motivation as a starting point for programming than creating a new
application from scratch.

An important design decision for tools in this space is how to deviate
from traditional spreadsheets like Microsoft Excel or Google Sheets.
Quilt and Glide use existing spreadsheet software as a backend,
providing maximum familiarity for users, and even compatibility with
existing spreadsheets. Gneiss has its own spreadsheet implementation
with additional features useful for building GUIs. Marmite provides a
live data view that resembles a spreadsheet, but programming is actually
done using a separate data flow pane rather than spreadsheet formulas.
(Marmite's approach led to some confusion in a user study, because users
expected behavior more similar to spreadsheets \autocite{wong2007}.)
Airtable deviates the furthest: although the user interface resembles a
spreadsheet, the underlying structure is a relational database with
typed columns. Wildcard's table is most similar to Airtable; the
structure of a relational table is most appropriate for most data in
websites, and we have not yet found a need for arbitrary untyped cells.
